\documentclass[a4paper]{report}
\input{../header.tex}
\author{Ryan Chan}
\title{ENGG2760 Probability for Engineers}

\thispagestyle{empty}
\addbibresource{reference.bib}

\usepackage{adjustbox}
\usepackage{centernot}


\begin{document}
\setlength\parindent{0pt}


\maketitle

\newpage

\begin{abstract}
	This is a note for \textbf{ENGG2760 - Probability for Engineers}. 
	
	Contents are adapted from the lecture notes of ENGG2760, prepared by \href{https://www.cse.cuhk.edu.hk/~farnia/}{Farzan Farnia}, as well as some online resources.

	This note is intended solely as a study aid. While I have done my best to ensure the accuracy of the content, I do not take responsibility for any errors or inaccuracies that may be present. Please use the material thoughtfully and at your own discretion.
	
	If you believe any part of this content infringes on copyright, feel free to contact me, and I will address it promptly.

	Mistakes might be found. So please feel free to point out any mistakes.

	This note also includes some extra content that isn't included in the syllabus of ENGG2760, but I found it somehow useful for overall understanding. They are mainly adapted from \href{https://ocw.mit.edu/RES-6-012S18}{MIT RES.6-012 Introduction to Probability}.
\end{abstract}

\newpage

\tableofcontents

\setlength{\parskip}{5pt}
\lec{1}{7}
%─────Appendix────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────
\appendix
\input{appendix.tex}

\end{document}