\documentclass[a4paper]{report}
\input{../header.tex}
\author{Ryan Chan}
\title{ENGG2780 Statistics for Engineers}

\thispagestyle{empty}
\addbibresource{reference.bib}

\usepackage{adjustbox}
\usepackage{centernot}


\begin{document}
\setlength\parindent{0pt}

\maketitle

\newpage

\begin{abstract}
	This is a note for \textbf{ENGG2780 - Statistics for Engineers}.
	
	Contents are adapted from the lecture notes of ENGG2780, prepared by \href{https://www.cse.cuhk.edu.hk/~sinnopan/}{Sinno Jialin Pan} and \href{https://andrejb.net}{Andrej Bogdanov}, as well as some online resources.

	This note is intended solely as a study aid. While I have done my best to ensure the accuracy of the content, I do not take responsibility for any errors or inaccuracies that may be present. Please use the material thoughtfully and at your own discretion.
	
	If you believe any part of this content infringes on copyright, feel free to contact me, and I will address it promptly.

	Mistakes might be found. So please feel free to point out any mistakes.
	
	This course heavily relies on prior knowledge of probability (which you can refer to in the notes I wrote for \href{https://www.ryanc.wtf/files/ENGG2760.pdf}{ENGG2760}). Therefore, before proceeding with this course, make sure you understand the foundation, as I will take them for granted.
\end{abstract}

\newpage

\tableofcontents

\setlength{\parskip}{5pt}
\lec{1}{9}
%─────Appendix────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────────
\appendix
\input{appendix.tex}

\end{document}